---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### 1. Load data from csv file 
    
```{r}
        setwd("C:/Users/Admin/Desktop/coursera/programmingAssignment_RR1/RepData_PeerAssessment1/activity")
        stepData <- read.csv("activity.csv")
        summary(stepData)
```

### 2. Process/transform data 
 
```{r} 
        # Convert date from factor to type date
        stepData$date <- as.Date(stepData$date)
        #Get all the unique dates
        uniqueDates <- unique(stepData$date)
        #keep a list of intervals
        uniqueIntervals <- unique(stepData$interval)
        
```        
        
#What is mean total of steps taken per day
### 1.Total number of steps per day         

```{r}
        #split the steps based on date
        stepsSplitPerDay <- split(stepData$steps, stepData$date)
        # sum of steps taken per day
        sumOfStepsPerDay <- sapply(stepsSplitPerDay, sum, na.rm = TRUE) 
        sumOfStepsPerDay
```        
### 2. Histogram of total steps per day

``` {r}
        
        plot(uniqueDates, sumOfStepsPerDay, 
             main = "Histogram of steps taken each day", 
             xlab = "Date october november", 
             ylab = "Frequency", 
             type = "h", lwd = 4, col = "blue")


```
### 3. Mean steps per day       

```{r}
        meanOfStepsPerDay <- sapply(stepsSplitPerDay, mean, na.rm = TRUE)
        meanDataFrame <- data.frame(uniqueDates, meanOfStepsPerDay, row.names = NULL)
        head(meanDataFrame)
```     
### 4. Median steps per day

``` {r}
        medianStepsPerDay <- sapply(stepsSplitPerDay, median, na.rm = TRUE)
        medianDataFrame <- data.frame(uniqueDates,medianStepsPerDay, row.names = NULL)
        head(medianDataFrame)
        
```
## Average daily acitivity pattern

```{r}
        intervalSplit <- split(stepData$steps, stepData$interval)  
        averageStepsPerInterval <- sapply(intervalSplit, mean, na.rm = TRUE)
        plot(uniqueIntervals, averageStepsPerInterval, main = "Activity pattern",
                        xlab = "Intervals", ylab = "steps per 5 minutes",
                        type = "l",col = "blue")
        
        
        maxIntervalDays <- max(averageStepsPerInterval, na.rm = TRUE)
        maxIndex <- as.numeric(which(averageStepsPerInterval == maxIntervalDays))
        
        maxInterval <- uniqueIntervals[maxIndex]
        abline(v=maxInterval, col="red", lwd=3)
```


## Imputing missing data values

### 1. Total number of missing values

```{r}
        isna <- is.na(stepData$steps)
        sum(isna)
        
```
### 2 Filling the missing values         

```{r}
         meanOfStepsPerDay[is.nan(meanOfStepsPerDay)] <- 0
         meanColumn <- rep(meanOfStepsPerDay, 288)
         rawSteps <- stepData$steps
         stepsNA <- is.na(rawSteps)
         rawSteps[stepsNA] <- meanColumn[stepsNA]
```
### 3. Create a new data set with NAs replaced by mean values

```{r}
        datNew <- stepData
        datNew$steps <- rawSteps
        #datNew
        
        stepsSplitNew <- split(datNew$steps, stepData$date)
        
        totalStepsPerDayNew <- sapply(stepsSplitNew, sum)
```  
### 4 Plot histogram for the new data and data with NAs

```{r}
        
        par(mfcol=c(2,1))
        # Plot the original histogram first
        plot(uniqueDates, sumOfStepsPerDay, main="Histogram of steps taken each day before imputing", 
                xlab="Date (October to November 2012)", ylab="Frequency", type="h", lwd=4, col="blue")
        # Plot the modified histogram after
        plot(uniqueDates, totalStepsPerDayNew, main="Histogram of steps taken each day after imputing", 
                xlab="Date (October to November 2012)", ylab="Frequency", type="h", lwd=4, col="blue")        

```
### Create a new factor variable in the steps data.

```{r}
        dates <- strptime(datNew$date,"%Y-%m-%d")
        wday <- dates$wday
        
        #classify the date as weekend or weekdays
        classifydays <- rep(0, length(wday)-1)
        
        #Fill classifydays with 1 if the date is a weekday
        classifydays[wday >= 1 & wday <=5] <- 1
        #Fill  classifydays with 2 if the date is a weekend
        classifydays[wday == 6 | wday == 0] <- 2
        
        #create a factor variable to our data set
        factorData <- factor(classifydays, levels = c(1,2), labels = c("Weekday", "Weekend"))
        
        datNew$dayOfWeek <- factorData 
        
        
        #split into weekdays and weekends
        daysweekday <- datNew[datNew$dayOfWeek == "Weekday",]
        daysweekend <- datNew[datNew$dayOfWeek == "Weekend",]
        
        head(datNew)
        
```
### Make a panel plot for the weekdays and weekend data.

```{r}

        dataSplitWeekEnd <- split(daysweekend$steps, daysweekend$interval)
        dataSplitWeekDay <- split(daysweekday$steps, daysweekday$interval)
        
        #get mean of each days steps
        
        meanStepsWeekEnd <- sapply(dataSplitWeekEnd, mean, na.rm = TRUE)
        meanStepWeekdays <- sapply(dataSplitWeekDay, mean, na.rm = TRUE)

        par(mfcol = c(2,1))
        
        plot(x = uniqueIntervals, y = meanStepWeekdays, type = "l", 
                main = "weekdays plot",xlab = "days", 
                ylab = "steps", lwd = 2, col = "blue")
        plot(x = uniqueIntervals, y = meanStepsWeekEnd, type = "l",
                main = "weekend plot", xlab = "days", 
                ylab = "steps", lwd = 2, col = "red")
```

